Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-06 ind_files/04_unification_discussion.tex
Add discussion for field eqns

>>>>

\section{Stochastic Field Equations for eco-evolutionary dynamics}

In chapter \ref{chap_infD_processes}, we formulated a `field equation' for the evolution of one-dimensional quantitative traits in populations. To recap, given a one-dimensional quantitative trait that takes values in some subset $\mathcal{T} \subset \mathbb{R}$, we defined the set $\mathcal{M}(\mathcal{T}) =  \left\{\sum_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}$, where each $\delta_{x_i}$ is a Dirac mass centered at $x_i$. We then formulated a model in which the population at time $t$ is characterized as a whole by a density distribution (`stochastic field') $\nu^{(t)}: \mathcal{M}(\mathcal{T}) \to [0,\infty)$ such that for any subset $A \subseteq \mathcal{T}$, the number of individuals that have trait value in $A$ is given by integrating $\nu^{(t)}$ over $A$. The change of this field is determined entirely by two functionals, $b(x|\nu)$ and $d(x|\nu)$ from $\mathcal{T}$ to $[0,\infty)$, that respectively describe the birth rate and death rate of individuals of type $x$ in a population $\nu$. Under the assumption that there exists a suitable system size parameter $K$ to go from $\mathcal{M}(
\mathcal{T})$ to $\mathcal{M}_{K}(\mathcal{T}) =  \left\{\sum_{i=1}^{n}\delta_{x_i}/K \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}$ , we found that $P(\phi,t)$, the probability that the population is described by the distribution $\phi \in \mathcal{M}_{K}(\mathcal{T})$ at time $t$, satisfies the `stochastic field equation':
\label{disc_functional_field_eqns}
\resizebox{\textwidth}{!}{$\displaystyle
\frac{\partial P}{\partial t}(\phi,t) = \int\limits_{\mathcal{T}}\left[-
\frac{\delta}{\delta\phi(x)} \{ \left(\phi(x) w(x|\phi) + \mu Q(x|\phi)\right) P(\phi,t) \} + \frac{1}{2K}\frac{\delta^2}{\delta\phi(x)^2}\{\left(\phi(x) \tau(x|\phi) + \mu Q(x|\phi)\right)P(\phi,t)\}\right]dx$}
where, as before, $w(x|\phi)$ and $\tau(x|\phi)$ are respectively the Malthusian fitness and per-capita turnover rate of type $x \in \mathcal{T}$ (now a continuous variable) in a population $\phi$. We then showed that this equation yields some well-known frameworks of quantitative genetics in the infinite population ($K \to \infty$) limit, thus illustrating consistency with known theories. If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes is true, then the exact same steps carried out in appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} for finite-dimensional processes will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} (with sums replaced by integrals and derivatives replaced by functional derivatives as appropriate) as the `fundamental theorems' for the evolution of quantitative traits. Indeed, \cite{week_white_2021} have recently proposed exactly the It\^o formula we would need via a heuristic It\^o multiplication table for $L^2(\mathbb{R}, m)$ valued stochastic processes with a.s. finite Hilbert-Schmidt norm and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by \cite{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like arcane gibberish to you, don't worry too much about the details - The essence is that \cite{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility over mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to certain SPDEs, \cite{week_white_2021} have arrived at exactly the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} from the `SPDE side'\footnote{If the well-known equivalence between SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) carries through to infinite-dimensions, it becomes an equivalence between SPDEs and functional Fokker-Planck equations} under certain additional assumptions required to ensure existence/uniqueness of solutions to the relevant SPDEs. Equation \eqref{disc_functional_field_eqns} can thus be argued to be the (functional) Fokker-Planck view of the stochastic processes for the evolutionary ecology of quantitative traits recently studied in \cite{week_white_2021}. Like \cite{week_white_2021}, I believe that rigorously establishing the precise criteria needed for general existence and uniqueness of solutions to equations of the form \eqref{disc_functional_field_eqns} (or the SPDE versions presented in \cite{week_white_2021}, which resemble our equation \eqref{functional_langevin}) presents an important mathematical direction for future work that may yield new insights into both the biology of quantitative traits and the mathematics of measure-valued stochastic processes.\\
More generally, the intimate relation between It\^o SDEs and Fokker-Planck equations has been extensively exploited in the applied mathematics literature \citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009} since different problems are more appropriately attacked from one side than the other. I too have repeatedly used this equivalence in chapter \ref{chap_BD} to arrive at our final equations. In the infinite-dimensional case, the functional Fokker-Planck equation \eqref{disc_functional_field_eqns} (the `physics' side) allows us to attack questions of phenotypic clustering and evolutionary branching via powerful tools such as Fourier analysis (\cite{rogers_demographic_2012, rogers_modes_2015}; See also appendix \ref{App_examples} for a general model-independent pedagogical derivation) that would have been significantly harder if one only used SPDEs. On the other hand, some questions are much more easily handled using tools from the `math' side of SPDEs such as infinitesimal generators and the martingale problem (see \cite{champagnat_unifying_2006} or \cite{week_white_2021}) rather than being forced to work with functional Fokker-Planck equations. Rigorously establishing a relation between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and SPDEs of the form \eqref{functional_langevin} (or those studied in \cite{week_white_2021}) for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between two alternative views of the same object. Importantly, the formalism I develop here in terms of functional Fokker-Planck equations likely does \emph{not} carry over to the study of higher dimensional quantitative traits (or populations which vary in two or more one-dimensional quantitative traits) because the probability density function $P(\phi,t)$ likely does not exist in two or more dimensions \citep{fleming_measure-valued_1979,champagnat_unifying_2006}. My understanding is that the evolution of quantitative traits is also very difficult to study in two or more dimensions from the SPDE side for largely the same reason, and overcoming this difficulty will likely need new approaches altogether.\\
To the best of my knowledge, a general formulation ot stochastic field equations from the functional Fokker-Planck side in the manner I have carried out here has not been done before. Similar equations have been formulated for some specific stochastic models of quantitative trait evolution \citep{rogers_demographic_2012,rogers_modes_2015} and population ecology of size-structured communities \citep{odwyer_integrative_2009}. Stochastic field equations are also known in mathematical neurobiology \citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014}, and have recently been proposed in a model of collective motion \citep{o_laighleis_minimal_2018}, suggesting that the study and formalization of such equations by pure mathematicians could have wide-spread applications. As is, such field equations are currently primarily used by physicists working in fields such as statistical field theory and quantum field theory, and are largely attacked through ingenious heuristic arguments and tools in a somewhat system-specific manner. Equation \eqref{disc_functional_field_eqns} opens up the stochastic evolution of quantitative traits to analysis using various powerful tools of physics and stochastic processes such as duality \citep{greenman_duality_2020}, the Fock space representation \citep{dodd_many-body_2009, del_razo_probabilistic_2022}, and the path integral formalism \citep{doi_second_1976, peliti_path_1985, dodd_many-body_2009, chow_path_2015, weber_master_2017}. Equation \eqref{disc_functional_field_eqns} also provides a more accessible alternative formulation that does not (explicitly) rely on tools from relatively `advanced' mathematical fields such as martingale theory and measure-valued processes that are typically used to study the evolution of quantitative traits in finite, fluctuating populations \citep{dawson_stochastic_1975,fleming_measure-valued_1979,champagnat_unifying_2006,etheridge_mathematical_2011,champagnat_polymorphic_2011,week_white_2021}, thus hopefully making the field more accessible to theorists without formal training in mathematical areas such as measure theory and stochastic processes.
